 

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al. 2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies, where dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. (*This sentence is too complex-- hard to understand*)

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].

Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@songTreeSurfaceTemperature2020].

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and greater $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. 
Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a]. At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007]

>>>>

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al. 2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies. Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. (*This sentence is too complex-- hard to understand*)

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].

Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

Leaf interiors are typically at maximum $RH$ [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018].

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves [Table 1]. However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. 
Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a]. At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007].
